# Introducing Linear Mixed-Effects Models

```{r setup, include=FALSE}
library("kableExtra")

## paste
.p <- paste0

## .fraction
.f <- function(x, y) {
  paste0("\\frac{", x, "}{", y, "}")
}

## y-bar
.yb1 <- function(x) {
  paste0("$\\bar{Y}_{", x, "}$")
}

.yb2 <- function(x) {
  paste0("\\bar{Y}_{", x, "}")
}

## subtraction term
.st <- function(x, y, bracket = NULL) {
  if (is.null(bracket)) {
    paste0(x, " - ", y)
  } else {
    paste0(bracket[1], x, " - ", y, bracket[2])
  }
}

.rb <- c("(", ")")
.dr <- c("\\displaystyle\\left(", "\\right)")
.ds <- c("\\displaystyle\\left[", "\\right]")
```

## Learning objectives

* express multi-level designs in mathematical format (DGP)
* analyze and interpret parameter estimates using `lme4::lmer()`
* simulate data from a multi-level design with a single predictor

In the exercises below, we'll be working with the `sleepstudy` data, a built in dataset from the lme4 package.  To get information about the data, after loading in the lme4 package, type `?sleepstudy` at the prompt.

## Modeling multi-level data

The `sleepstudy` data is included as a built-in dataset in the `lme4` package for R [@Bates_et_al_2015]. Some of the ideas presented in this section come from @McElreath_2020 [Statistical Rethinking](https://xcelab.net/rm/statistical-rethinking/) textbook as well as from Tristan Mahr's [blog post on partial pooling](https://www.tjmahr.com/plotting-partial-pooling-in-mixed-effects-models/).

Let's start by looking at the documentation for the `sleepstudy` dataset. After loading the **`lme4`** package, you can access the documentation by typing `?sleepstudy` in the console.

```
sleepstudy                package:lme4                 R Documentation

Reaction times in a sleep deprivation study

Description:

     The average reaction time per day (in milliseconds) for subjects
     in a sleep deprivation study.

     Days 0-1 were adaptation and training (T1/T2), day 2 was baseline
     (B); sleep deprivation started after day 2.

Format:

     A data frame with 180 observations on the following 3 variables.

     ‘Reaction’ Average reaction time (ms)

     ‘Days’ Number of days of sleep deprivation

     ‘Subject’ Subject number on which the observation was made.

Details:

     These data are from the study described in Belenky et al.  (2003),
     for the most sleep-deprived group (3 hours time-in-bed) and for
     the first 10 days of the study, up to the recovery period.  The
     original study analyzed speed (1/(reaction time)) and treated day
     as a categorical rather than a continuous predictor.

References:

     Gregory Belenky, Nancy J. Wesensten, David R. Thorne, Maria L.
     Thomas, Helen C. Sing, Daniel P. Redmond, Michael B. Russo and
     Thomas J. Balkin (2003) Patterns of performance degradation and
     restoration during sleep restriction and subsequent recovery: a
     sleep dose-response study. _Journal of Sleep Research_ *12*, 1-12.
```

Note that these data meet our definition of multilevel data due to multistage sampling: Participants are sampled from a population, and then their mean reaction times are measured on each of the 10 days in the study. Even though we don't have the trial-by-trial information used to calculate the means, this is still multilevel data, because of the multiple observations on the response variable (reaction time) across days for each subject.

Multi-level data of this type is extremely common in psychology. However, if you open up most introductory statistics textbooks and read through to the end, you'll learn about t-tests, repeated-measures and mixed-model ANOVA, and multiple regression, but none of the techniques that you learn would be applicable to this type of data. It suggests a serious gap in training. Because most data is multi-level, multi-level modeling should be taught as the **default** approach, not as an advanced approach that is far beyond the reach of undergraduates. All of the other techniques---t-test, ANOVA, multiple regression---can be seen as special cases of the more general multi-level approach. 

Let's consider the `sleepstudy` data. The dataset contains eighteen participants from the three-hour sleep condition. Each day, over 10 days, participants performed a ten-minute "psychomotor vigilance test" where they had to monitor a display and press a button as quickly as possible each time a stimulus appeared. The dependent measure in the dataset is the participant's average response time (RT) on the task for that day.

It is always good to start by looking at the data.  Here is data for a single subject. 

```{r one-subject, message = FALSE, fig.cap = "*Data from a single subject in Belenky et al. (2003)*", fig.width = 4, fig.height = 3, out.width = "50%"}
library("lme4")
library("tidyverse")

just_308 <- sleepstudy %>%
  filter(Subject == "308")

ggplot(just_308, aes(x = Days, y = Reaction)) +
  geom_point() +
  scale_x_continuous(breaks = 0:9)
```

<div class="try">

### Exercise {-}

Use ggplot to recreate the plot below, which shows data for all 18 subjects.

```{r plot-solution0, echo = FALSE, message = FALSE, fig.cap = "*Data from Belenky et al. (2003)*"}
ggplot(sleepstudy, aes(x = Days, y = Reaction)) +
  geom_point() +
  scale_x_continuous(breaks = 0:9) +
  facet_wrap(~Subject)
```

It looks like RT is increasing with each additional day of sleep deprivation, starting from day 2 and increasing until day 10.

`r hide("Hint, please")`

Just above, you were given the code to make a plot for a single participant. Adapt this code to show all of the participants by getting rid of the `filter()` statement and adding a *`ggplot2`* function that starts with `facet_`.

`r unhide()`

`r hide("Show me")`

Same as above, except you just add one line: `facet_wrap(~Subject)`

```{r plot-solution, eval=FALSE}
ggplot(sleepstudy, aes(x = Days, y = Reaction)) +
  geom_point() +
  scale_x_continuous(breaks = 0:9) +
  facet_wrap(~Subject)
```

`r unhide()`

</div>

## How to model these data?

This is how @Belenky_et_al_2003 describe their study (p. 2 of the PDF):

> The first 3 days (T1, T2 and B) were adaptation and training (T1 and T2) and baseline (B) and subjects were required to be in bed from 23:00 to 07:00 h [8 h required time in bed (TIB)]. On the third day (B), baseline measures were taken. Beginning on the fourth day and continuing for a total of 7 days (E1–E7) subjects were in one of four sleep conditions [9 h required TIB (22:00–07:00 h), 7 h required TIB (24:00–07:00 h), 5 h required TIB (02:00–07:00 h), or 3 h required TIB (04:00–07:00 h)], effectively one sleep augmentation condition, and three sleep restriction conditions.

There were seven nights of sleep restriction, with the first night of restriction occurring after the third day. The first two days, coded as `0`, `1`, were adaptation and training. The day coded as `2`, where the baseline measurement was taken, should be the place where we start our analysis. If we include the days `0` and `1` in our analysis, this might bias our results, since any changes in performance during the first two days have to do with training, not sleep restriction.

<div class="try">

### Exercise {-}

Remove from the dataset observations where `Days` is coded `0` or `1`, and then make a new variable `days_deprived` from the `Days` variable so that the sequence starts at day 2, with day 2 being re-coded as day 0, day 3 as day 1, day 4 as day 2, etc. This new variable now tracks the number of days of sleep deprivation. Store the new table as `sleep2`.

`r hide("Solution")`

```{r recode-vars}
sleep2 <- sleepstudy %>%
  filter(Days >= 2L) %>%
  mutate(days_deprived = Days - 2L)
```

It is always a good idea to double check that the code works as intended. First, look at it:

```{r look-at-the-data}
head(sleep2)
```

And check that `Days` and `days_deprived` match up.

```{r recode-test}
sleep2 %>%
  count(days_deprived, Days)
```

Looks good. Note that the variable `n` in generated by `count()` and tells you how many rows there are for each unique combination of `Days` and `days_deprived`. In this case, there were 18, one row for each participant.

`r unhide()`

</div>

Now let's re-plot the data looking at just these eight data points from Day 0 to Day 7. We've just copied the code from above, substituting `sleep2` for `sleepstudy` and using `days_deprived` for our `x` variable.

```{r plot-solution2, message = FALSE, fig.cap = "*Data from Belenky et al. (2003), showing reaction time at baseline (0) and after each day of sleep deprivation.*"}
ggplot(sleep2, aes(x = days_deprived, y = Reaction)) +
  geom_point() +
  scale_x_continuous(breaks = 0:7) +
  facet_wrap(~Subject) +
  labs(y = "Reaction Time", x = "Days deprived of sleep (0 = baseline)")
```

Take a moment to think about how me might model the relationship between `days_deprived` and `Reaction`. Does reaction time increase or decrease with increasing sleep deprivation? Is the relationship roughly stable or does it change with time?

With only one exception (subject 335) it looks like reaction time increases with each additional day of sleep deprivation. It looks like we could fit a line to each participant's data. Recall the general equation for a line is of the form **y = y-intercept + slope $\times$ x**. In regression, the we usually express a linear relationship with the formula

$$Y = \beta_0 + \beta_1 X$$

where $\beta_0$ is the y-intercept and $\beta_1$ is the slope, parameters whose values we estimate from the data.

The lines will all differ in intercept (mean RT at day zero, before the sleep deprivation began) and slope (the change in RT with each additional day of sleep deprivation). But should we fit the same line to everyone? Or a totally different line for each subject? Or something somewhere in between? 

Let's start by considering three different approaches we might take. Following McElreath, we will distinguish these approaches by calling them **complete pooling**, **no pooling**, and **partial pooling**.

### Complete pooling: One size fits all

The **complete pooling** approach is a "one-size-fits-all" model: it estimates a single intercept and slope for the entire dataset, ignoring the fact that different subjects might vary in their intercepts or slopes. If that sounds like a bad approach, it is; but you know this because you've already visualized the data and noted that the pattern for each participant would seem to require different y-intercept and slope values.

Fitting one line is called the "complete pooling" approach because we pool together data from all subjects to get single estimates for an overall intercept and slope. The GLM for this approach is simply

$$Y_{sd} = \beta_0 + \beta_1 X_{sd} + e_{sd}$$

$$e_{sd} \sim N\left(0, \sigma^2\right)$$

where $Y_{sd}$ is the mean RT for subject $s$ on day $d$, $X_{sd}$ is the value of `Day` associated with that case (0-7), and $e_{sd}$ is the error.

We would fit such a model in R using the `lm()` function, e.g.:

```{r complete-pooling-lm}
cp_model <- lm(Reaction ~ days_deprived, sleep2)

summary(cp_model)
```

According to this model, the predicted mean response time on Day 0 is about `r coef(cp_model)[1] %>% round()` milliseconds, with an increase of about `r coef(cp_model)[2] %>% round()` milliseconds per day of deprivation, on average.  We can't trust the standard errors for our regression coefficients, however, because we are assuming that all of the observations are independent (technically, that the residuals are). However, we can be pretty sure this is a bad assumption.

Let's add the model predictions to the graph that we created above. We can use `geom_abline()` to do so, specifying the intercept and slope for the line using the regression coefficients from the model fit, `coef(cp_model)`, which returns a two-element vector with intercept and slope, respectively.

```{r coef-example}
coef(cp_model)
```

```{r cp-model-plot, fig.cap="Data plotted against predictions from the complete pooling model."}
ggplot(sleep2, aes(x = days_deprived, y = Reaction)) +
  geom_abline(intercept = coef(cp_model)[1],
              slope = coef(cp_model)[2],
              color = 'blue') +
  geom_point() +
  scale_x_continuous(breaks = 0:7) +
  facet_wrap(~Subject) +
  labs(y = "Reaction Time", x = "Days deprived of sleep (0 = baseline)")
```

The model fits the data badly. We need a different approach.

### No pooling

Pooling all the information to get just one intercept and one slope estimate is inappropriate. Another approach would be to fit separate lines for each participant. This means that the estimates for each participant will be completely uninformed by the estimates for the other participants. In other words, we can separately estimate 18 individual intercept/slope pairs.

This model could be implemented in two ways: (1) by running separate regressions for each participant or (2) by running fixed-effects regression. We'll do the latter, so that everything is in one big model. We know how to do this already: we add in dummy codes for the `Subject` factor. We have 18 levels of this factor, so we'd need 17 dummy codes. Fortunately, R saves us from the trouble of creating the 17 variables we would need by hand. All we need to do is include `Subject` as a predictor in the model, and interact this categorical predictor with `days_deprived` to allow intercepts and slopes to vary.

<div class="danger">

The variable `Subject` in the `sleep2` dataset is nominal. We just use numbers as labels to preserve anonymity, without intending to imply that Subject 310 is one point better than Subject 309 and two points better than 308. Make sure that you define it as a factor so that it is not included as a continuous variable!

We can test whether something is a factor in various ways. One is to use `summary()` on the table.

```{r summary}
sleep2 %>% summary()
```

Here you can see that it is not treated as a number because rather than giving you distributional information (means, etc.) it tells you how many observations there are at each level.

You can also test it directly:

```{r is-factor}
sleep2 %>% pull(Subject) %>% is.factor()
```

If something is not a factor, you can make it one be re-defining it using the `factor()` function.

</div>

```{r hidden-no-pooling, echo = FALSE}
.np <- lm(Reaction ~ days_deprived + Subject + days_deprived:Subject, sleep2)
```


```{r lm2-no-pooling}
np_model <- lm(Reaction ~ days_deprived + Subject + days_deprived:Subject,
               data = sleep2)

summary(np_model)
```

What this model has done is take one subject to be the baseline (specifically, subject 308), and represent each subject in terms of offsets from that baseline. You saw this already when we talked about [continuous-by-categorical interactions](05-interactions-course-notes.html#cont-by-cat).

<div class="try">

Answer these questions (to three decimal places):

* What is the intercept for subject 308? `r fitb(coef(.np)["(Intercept)"], width = 8, num = TRUE, tol = .001)`
* What is the slope for subject 308? `r fitb(coef(.np)["days_deprived"], num = TRUE, width = 8, tol = .001)`
* What is the intercept for subject 335? `r fitb(coef(.np)["(Intercept)"] + coef(.np)["Subject335"], width = 8, num = TRUE, tol = .001)`
* What is the slope for subject 335? `r fitb(coef(.np)["days_deprived"] + coef(.np)["days_deprived:Subject335"], num = TRUE, width = 8, tol = .001)`

`r hide("Answers and Explanation")`

The baseline subject is 308; the default in R is to sort the levels of the factor alphabetically and chooses the first one as the baseline. This means that the intercept and slope for 308 are given by `(Intercept)` and `days_deprived` respectively, because all of the other 17 dummy variables will be zero for subject 308.

All of the regression coefficients for the other subjects are represented as *offsets* from this baseline subject. If we want to calculate the intercept and slope for a given subject, we just add in the corresponding offsets. So, the answers are

* intercept for 308: <u>`r coef(.np)["(Intercept)"] %>% round(3)`</u>
* slope for 308: `r coef(.np)["days_deprived"] %>% round(3)`

* intercept for 335: `(Intercept)` + `Subject335` = `r .c1 <- coef(.np)["(Intercept)"] %>% round(3); .c1` + `r .c2 <- coef(.np)["Subject335"] %>% round(3); .c2` = <u>`r round(.c1 + .c2, 3)`</u>
* slope for 335: `days_deprived` + `days_deprived:Subject335` = `r .c3 <- coef(.np)["days_deprived"] %>% round(3); .c3` + `r .c4 <- coef(.np)["days_deprived:Subject335"] %>% round(3); .c4` = <u>`r round(.c3 + .c4, 3)`</u>

`r unhide()`

</div>

In the "no pooling" model, there is no *overall* population intercept and slope that is being estimated; in this case, `(Intercept)` and `days_deprived` are estimates of the intercept and slope for subject 308, which was (arbitrarily) chosen as the baseline subject. To get population estimates, we could introduce a second stage of analysis where we calculate means of the individual intercepts and slopes. Let's use the model estimates to calculate the intercepts and slopes for each subject.

```{r no-pooling-second-stage}
all_intercepts <- c(coef(np_model)["(Intercept)"],
                    coef(np_model)[3:19] + coef(np_model)["(Intercept)"])

all_slopes  <- c(coef(np_model)["days_deprived"],
                 coef(np_model)[20:36] + coef(np_model)["days_deprived"])

ids <- sleep2 %>% pull(Subject) %>% levels() %>% factor()

# make a tibble with the data extracted above
np_coef <- tibble(Subject = ids,
                  intercept = all_intercepts,
                  slope = all_slopes)

np_coef
```

Let's see how well this model fits our data.

```{r np-model-fits, fig.cap = "Data plotted against fits from the no-pooling approach."}
ggplot(sleep2, aes(x = days_deprived, y = Reaction)) +
  geom_abline(data = np_coef,
              mapping = aes(intercept = intercept,
                            slope = slope),
              color = 'blue') +
  geom_point() +
  scale_x_continuous(breaks = 0:7) +
  facet_wrap(~Subject) +
  labs(y = "Reaction Time", x = "Days deprived of sleep (0 = baseline)")
```

This is much better than the complete pooling model. If we want to test the null hypothesis that the fixed slope is zero, we could do using a one-sample test.

```{r np-inference-slope-hidden, include=FALSE}
.np1 <- np_coef %>% pull(slope) %>% t.test()

```

```{r np-inference-slope}
np_coef %>% pull(slope) %>% t.test()
```

This tells us that the mean slope of
`r np_coef %>% pull(slope) %>% mean() %>% round(3)`
is significantly different from zero,
`r sprintf("t(%d) = %0.3f", .np1$parameter, .np1$statistic)`, $p < .001$.

### Partial pooling using mixed-effects models

Neither the complete or no-pooling approach is satisfactory. It would be desirable to improve our estimates for individual participants by taking advantage of what we know about the other participants. This will help us better distinguish signal from error for each participant and improve generalization to the population. As the web app below will show, this becomes particularly important when we have unbalanced or missing data.

In the no-pooling model, we treated `Subject` as a `r glossary("fixed factor", display = "**fixed factor**")`. Each pair of intercept and slope estimates is determined by that subject's data alone. However, we are not interested in these 18 subjects in and of themselves; rather, we are interested in them as examples drawn from a larger population of potential subjects. This subjects-as-fixed-effects approach is suboptimal if your goal is to generalize to new participants in the population of interest. 

Partial pooling happens when you treat a factor as a random instead of fixed in your analysis. A `r glossary("random factor", display = "**random factor**")` is a factor whose levels are considered to represent a proper subset of all the levels in the population. Usually, you treat a factor as random when the levels you have in the data are the result of sampling, and you want to generalize beyond those levels. In this case, we have eighteen unique subjects and thus, eighteen levels of the `Subject` factor, and would like to say something general about effects of sleep deprivation on the population of potential subjects.

Just as there are two types of factors---fixed and random---there are 

A way to include random factors in your analysis is to use a `r glossary("linear mixed-effects model")`. 

estimates at each level of the factor (i.e., for each subject) become informed by information from other levels (i.e., other subjects). Rather than estimating the intercept and slope for each participant without considering the estimates for other subjects, the model estimates values for the population, and pulls the estimates for individual subjects toward those values, a statistical phenomenon known as `shrinkage`. 

The population estimates are informed by the individual estimates, but each individual estimate is "pulled" toward the population value.

The multilevel model is below. It is important that you understand the math and what it means. It looks complicated at first, but there's really nothing below that you haven't seen before. We'll explain everything step by step.

*Level 1:*

\begin{equation}
Y_{sd} = \beta_{0s} + \beta_{1s} X_{sd} + e_{sd}
\end{equation}

*Level 2:*

\begin{equation}
\beta_{0s} = \gamma_{0} + S_{0s}
\end{equation}

\begin{equation}
\beta_{1s} = \gamma_{1} + S_{1s}
\end{equation}

*Variance Components:*

\begin{equation}
 \langle S_{0s}, S_{1s} \rangle \sim N\left(\langle 0, 0 \rangle, \mathbf{\Sigma}\right) 
\end{equation}

\begin{equation}
\mathbf{\Sigma} = \left(\begin{array}{cc}{\tau_{00}}^2 & \rho\tau_{00}\tau_{11} \\
         \rho\tau_{00}\tau_{11} & {\tau_{11}}^2 \\
         \end{array}\right) 
\end{equation}

\begin{equation}
e_{sd} \sim N\left(0, \sigma^2\right)
\end{equation}

In case you get lost, here's a table with an explanation for all of the variables in the set of equations above.

| Variable                 | Type      | Description                                                           |
|:-------------------------|:----------|:----------------------------------------------------------------------|
| $Y_{sd}$               | observed | Value of `Reaction` for subject $s$ on day $d$                    |
| $X_{sd}$               | observed | Value of `Day` (0-9) for subject $s$ on day $d$                  |
| $\beta_{0s}$           | derived  | level 1 intercept parameter                                           |
| $\beta_{1s}$           | derived  | level 1 slope parameter                                               |
| $e_{sd}$               | derived  | Residual ($Y_{sd}$ - $\hat{Y}_{sd}$) for subject $s$, day $d$ |
| $\gamma_0$             | fixed    | Grand intercept ("gamma")                                             |
| $\gamma_1$             | fixed    | Grand slope ("gamma")                                                 |
| $S_{0s}$               | derived  | Random intercept (offset) for subject $s$                           |
| $S_{1s}$               | derived  | Random slope (offset) for subject $s$                               |
| $\mathbf{\Sigma}$      | random   | Variance-covariance matrix                                            |
| ${\tau_{00}}^2$        | random   | Variance of random intercepts                                         |
| $\rho$                 | random   | Random correlation between intercepts and slopes                      |
| ${\tau_{11}}^2$        | random   | Variance of random slopes                                             |
| $\sigma^2$             | random   | Error variance                                                        |

Note the "Status" column of the table contains values *fixed*, *random*, and *derived*. Although *fixed* and *random* are standard terms, *derived* is not; I have introduced it to help you think about what these different variables mean in the context of the model and to help you distinguish variables that are directly estimated from variables that are not.

Let's begin with the Level 1 equation of our model, which represents the general relationship between the predictors and response variable. It captures the functional form of the main relationship between reaction time \(Y_{sd}\) and sleep deprivation \(X_{sd}\): a straight line with intercept \(\beta_{0s}\) and slope \(\beta_{1s}\). Now \(\beta_{0s}\) and \(\beta_{1s}\) makes it look like the complete pooling model, where we estimated a single intercept and single slope for the entire dataset; however, we're not actually estimating these directly. Instead, we're going to think of \(\beta_{0s}\) and \(\beta_{1s}\) as derived: they are wholly defined by variables at Level 2 of the model.

Level 2 of the model, defined by two equations, represents relationships at the participant level. Here, we define the intercept $\beta_{0s}$ in terms of a fixed effect \(\gamma_0\) and a *random intercept* \(S_{0s}\); likewise, we define the slope $\beta_{1s}$ in terms of a fixed slope \(\gamma_1\) and a *random slope* \(S_{1s}\).

The final equations represent the *Variance Components* of the model.  We'll get into this more in detail below.

Let's substitute the Level 2 equations into Level 1 to see the advantages of representing things in the multilevel way.

\begin{equation}
Y_{sd} = \gamma_{0} + S_{0s} + \left(\gamma_{1} + S_{1s}\right) X_{sd} + e_{sd}
\end{equation}

While this "combined" formula syntax is easy enough to understand in this particular case, the multilevel form more clearly allows us to see the functional form of the model: a straight line. We could easily change the functional form to, for instance, capture non-linear trends:

$$Y_{sd} = \beta_{0s} + \beta_{1s} X_{sd} + \beta_{2s} X_{sd}^2 + e_{sd}$$

This functional form gets obscured in the combined syntax.  The multilevel syntax also makes it easy to see which terms go with the intercept and which terms go with the slope. Also, as designs get more compilicated---for example, if we were to assign participants to different experimental conditions, thus introducing a further predictors at Level 2---the combined equations get harder and harder to parse and reason about.

Fixed effect parameters like \(\gamma_0\) and \(\gamma_1\) are estimated from the data, and reflect stable properties of the population. In this example, $\gamma_0$ is the population intercept and $\gamma_1$ is the population slope. You can think of these **fixed-effects parameters** as representing the **average intercept and slope** in the population. These are "fixed" in the sense that we assume that they reflect the true underlying values in the population; they are not assumed to vary from sample to sample.  These fixed effects parameters are often of prime theoretical interest; we want to measure them and their standard errors in a manner that is as unbiased and precise as the data allow. In experimental settings they are often the targets of hypothesis tests.

Random effects like \(S_{0i}\) and \(S_{1i}\) allow intercepts and slopes (respectively) to vary over subjects.  These random effects are *offsets*: deviations from the population 'grand mean' values. Some subjects will just be slower responders than others, such that they will have a higher intercept (mean RT) on day 0 than the population's estimated value of \(\hat{\gamma_0}\). These slower-than-average subjects will have positive \(S_{0i}\) values; faster-than-average subjects will have negative \(S_{0i}\) values. Likewise, some subjects will show stronger effects of sleep deprivation (steeper slope) than the estimated population effect, \(\hat{\gamma_1}\), which implies a positive offset \(S_{1s}\), while others may show weaker effects or close to none at all (negative offset).

Each participant can be represented as a vector pair \(\langle S_{0i}, S_{1i} \rangle\).  If the subjects in our sample comprised the entire population, we would be justified in treating them as fixed and estimating their values, as in the "no-pooling" approach above. This is not the case here. In recognition of the fact they are sampled, we are going to treat subjects as a random factor rather than a fixed factor. Instead of estimate the values for the subjects we happened to pick, we will estimate **the covariance matrix that represents the  bivariate distribution from which these pairs of values are drawn**. By doing this, we allow the subjects in the sample to inform us about characteristics of the population.

### The variance-covariance matrix

\begin{equation}
 \langle S_{0s}, S_{1s} \rangle \sim N\left(\langle 0, 0 \rangle, \mathbf{\Sigma}\right) 
\end{equation}

\begin{equation}
\mathbf{\Sigma} = \left(\begin{array}{cc}{\tau_{00}}^2 & \rho\tau_{00}\tau_{11} \\
         \rho\tau_{00}\tau_{11} & {\tau_{11}}^2 \\
         \end{array}\right) 
\end{equation}

Equations in the *Variance Components* characterize our estimates of variability. The first equation states our assumption that the random intercept / random slope pairs \(\langle S_{0s}, S_{1s} \rangle\) are drawn from a bivariate normal distribution centered at the origin \(\langle 0, 0 \rangle\) with variance-covariance matrix \(\mathbf{\Sigma}\).

The variance-covariance matrix is key: it determines the probability of drawing random effect pairs \(\langle S_{0s}, S_{1s} \rangle\) from the population.  You have seen these before, in the [chapter on correlation and regression](03-bivariate-simulation.html). The covariance matrix is always a square matrix (equal numbers of columns and rows).  On the main diagonal (upper left and bottom right cells) it has random effect variances \({\tau_{00}}^2\) and \({\tau_{11}}^2\). \({\tau_{00}}^2\) is the random intercept variance, which captures how much subjects vary in their mean response time on Day 0, before any sleep deprivation. \({\tau_{11}}^2\) is the random slope variance, which captures how much subjects vary in their susceptibility to the effects of sleep deprivation.

The cells in the off-diagonal contain covariances, but this information is represented redundantly in the matrix; the lower left element is identical to the upper right element; both capture the covariance between random intercepts and slopes, as expressed by \(\rho\tau_{00}\tau_{11}\). In this equation \(\rho\) is the correlation between the intercept and slope. So, all the information in the matrix can be captured by just three parameters: \(\tau_{00}\), \(\tau_{11}\), and \(\rho\).

## Estimating the model parameters

To estimate parameters, we are going to use the `lmer()` function of the lme4 package (Bates, Mächler, Bolker, & Walker, 2015).  The basic syntax of `lmer()` is

```
lmer(formula, data, ...)
```

where `formula` expresses the structure of the underlying model in a compact format and `data` is the data frame where the variables mentioned in the formula can be found.

The general format of the model formula for N fixed effects (`fix`) and K random effects (`ran`) is

`DV ~ fix1 + fix2 + ... + fixN + (ran1 + ran2 + ... + ranK | random_factor1)`

Interactions between factors A and B can be specified using either `A * B` (interaction and main effects) or `A:B` (just the interaction).

A key difference from standard R model syntax is the presence of a random effect term, which is enclosed in parentheses, e.g., `(ran1 + ran2 + ... + ranK | random_factor)`.  Each bracketed expression represents random effects associated with a single random factor. You can have more than one random effects term in a single formula, as we will see when we talk about crossed random factors. You should think of the random effects terms as providing `lmer()` with **instructions on how to construct variance-covariance matrices**.

On the left side of the bar `|` you put the effects you want to allow to vary over the levels of the random factor named on the right side. Usually, the right-side variable is one whose values uniquely identify individual subjects (e.g., `subject_id`).

Consider the following possible model formulas for the `sleep2` data and the variance-covariance matrices they construct.

|   | model                        | syntax                                      |
|---+------------------------------+---------------------------------------------|
| 1 | random intercepts only       | `Reaction ~ days_deprived + (1 | Subject)`         |
| 2 | random intercepts and slopes | `Reaction ~ days_deprived + (1 + days_deprived | Subject)`  |
| 3 | model 2 alternate syntax     | `Reaction ~ days_deprived + (days_deprived | Subject)`      |
| 4 | random slopes only           | `Reaction ~ days_deprived + (0 + days_deprived | Subject)`  |
| 5 | model 2 + zero-covariances   | `Reaction ~ days_deprived + (days_deprived || Subject)` |

*Model 1:*

\begin{equation*}
  \mathbf{\Sigma} = \left(
  \begin{array}{cc}
    {\tau_{00}}^2 & 0 \\
                0 & 0 \\
  \end{array}\right) 
\end{equation*}

*Models 2 and 3:*

\begin{equation*}
  \mathbf{\Sigma} = \left(
  \begin{array}{cc}
             {\tau_{00}}^2 & \rho\tau_{00}\tau_{11} \\
    \rho\tau_{00}\tau_{11} &          {\tau_{11}}^2 \\
  \end{array}\right) 
\end{equation*}

*Model 4:*

\begin{equation*}
  \mathbf{\Sigma} = \left(
  \begin{array}{cc}
    0 &             0 \\
    0 & {\tau_{11}}^2 \\
  \end{array}\right) 
\end{equation*}

*Model 5:*

\begin{equation*}
  \mathbf{\Sigma} = \left(
  \begin{array}{cc}
    {\tau_{00}}^2 &             0 \\
                0 & {\tau_{11}}^2 \\
  \end{array}\right) 
\end{equation*}

The most reasonable model for these data is Model 2, so we'll stick with that.

Let's fit the model, storing the result in object `pp_mod`.

```{r mod}
pp_mod <- lmer(Reaction ~ days_deprived + (days_deprived | Subject), sleep2)

summary(pp_mod)
```

Before discussing how to interpret the output, let's first plot the data against our model predictions. We can get model predictions using the `predict()` function (see `?predict.merMod` for information about use with mixed-effects models).

First, create a new data frame with predictor values for `Subject` and `days_deprived`.

```{r pp-newdata}
newdata <- crossing(
  Subject = sleep2 %>% pull(Subject) %>% levels() %>% factor(),
  days_deprived = 0:7)

newdata
```

Then, run this through `predict()`. Typically we will add the prediction in as a new variable in the data frame of new data, giving it the same name as our DV (`Reaction`).

```{r pred-vals}
newdata2 <- newdata %>%
  mutate(Reaction = predict(pp_mod, newdata))
```

Now we are ready to plot.

```{r pp-plot, fig.cap = "Data plotted against predictions from a partial pooling approach."}
ggplot(sleep2, aes(x = days_deprived, y = Reaction)) +
  geom_line(data = newdata2,
            color = 'blue') +
  geom_point() +
  scale_x_continuous(breaks = 0:7) +
  facet_wrap(~Subject) +
  labs(y = "Reaction Time", x = "Days deprived of sleep (0 = baseline)")
```

## Interpreting `lmer()` output and extracting estimates

The call to `lmer()` returns a fitted model object of class "lmerMod". To find out more about the `lmerMod` class, which is in turn a specialized version of the `merMod` class, see `?lmerMod-class`.

```{r hid-parse-output, include = FALSE}
.mod_out <- capture.output(summary(pp_mod))

.headings <- c("^Random effects:$",
               "^Fixed effects:$",
               "^Correlation of Fixed Effects:$")

.head_ix <- map_int(.headings, grep, x = .mod_out)
```

### Fixed effects

The section of the output called `Fixed effects:` should look familiar; it is similar to what you would see in the output for a simple linear model fit by `lm()`.

```{r fixed-effects, echo=FALSE}
cat(.mod_out[.head_ix[2]:(.head_ix[3] - 2L)], sep = "\n")
```

This indicates that the estimated mean reaction time for participants at Day 0 was about
`r round(fixef(pp_mod) %>% pluck(1), 0)` milliseconds, 
with each day of sleep deprivation adding an additional 
`r round(fixef(pp_mod) %>% pluck(2), 0)` milliseconds 
to the response time, on average.

If we need to get the fixed effects from the model, we can extract them using `fixef()`.

```{r fixef}
fixef(pp_mod)
```

The standard errors give us estimates of the variability for these parameters due to sampling error.  You could use these to calculate the \(t\)-values or derive confidence intervals. Extract them using `vcov(pp_mod)` which gives a variance-covariance matrix (*not* the one associated with the random effects), pull out the diagonal using `diag()` and then take the square root using `sqrt()`.

```{r vcov}
sqrt(diag(vcov(pp_mod)))

# OR, equivalently using pipes:
# vcov(pp_mod) %>% diag() %>% sqrt()
```

Note that these \(t\) values do not appear with \(p\) values, as is customary in simpler modeling frameworks. There are multiple approaches for getting \(p\) values from mixed-effects models, with advantages and disadvantages to each; see @Luke_2017 for a survey of options. The \(t\) values do not appear with degrees of freedom, because the degrees of freedom in a mixed-effects model are not well-defined. Often people will treat them as Wald \(z\) values, i.e., as observations from the standard normal distribution. Because the \(t\) distribution asymptotes the standard normal distribution as the number of observations goes to infinity, this "t-as-z" practice is legitimate if you have a large enough set of observations.

To calculate the Wald \(z\) values, just divide the fixed effect estimate by its standard error:

```{r t-as-z}
tvals <- fixef(pp_mod) / sqrt(diag(vcov(pp_mod)))

tvals
```

You can get the associated \(p\)-values using the following formula:

```{r p-values}
2 * (1 - pnorm(abs(tvals)))
```

This gives us strong evidence against the null hypothesis \(H_0: \gamma_1 = 0\). Sleep deprivation does appear to increase response time.

You can get confidence intervals for the estimates using `confint()` (this technique uses the *parametric bootstrap*).  `confint()` is a generic function, so to get help on this function, use `?confint.merMod`.

```{r conf-int, cache=TRUE}
confint(pp_mod)
```

### Random effects

```{r ranfx, echo=FALSE}
cat(.mod_out[.head_ix[1]:(.head_ix[2] - 2L)], sep = "\n")
```

The random effects part of the `summary()` output is less familiar. What you find here is a table with information about the variance components: the variance-covariance matrix (or matrices, if you have multiple random factors) and the residual variance.

Let's start with the `Residual` line. This tells us that the residual variance, \(\sigma^2\), was estimated at about
`r round(sigma(pp_mod)^2, 2)`. The value in the next column, 
`r round(sigma(pp_mod), 3)`, is just the standard deviation, \(\sigma\), which is the square root of the variance.

We extract the residual standard deviation using the `sigma()` function.

```{r sigma}
sigma(pp_mod) # residual
```

The two lines above the `Residual` line give us information about the variance-covariance matrix for the `Subject` random factor.

```{r varcorr, echo = FALSE}
cat(.mod_out[(.head_ix[1] + 1L):(.head_ix[2] - 4L)], sep = "\n")
```

The values in the `Variance` column gives us the main diagonal of the matrix, and the `Std.Dev.` values are just the square roots of these values. The `Corr` column tells us the correlation between the intercept and slope.

We can extract these values from the fitted object `pp_mod` using the `VarCorr()` function. This returns a named list, with one element for each random factor. We have `Subject` as our only random factor, so the list will just be of length 1.

```{r get-varcorr}
# variance-covariance matrix for random factor Subject
VarCorr(pp_mod)[["Subject"]] # equivalently: VarCorr(pp_mod)[[1]]
```

The first few lines are a printout of the variance covariance matrix. You can see the variances in the main diagonal. We can get these with:

```{r varcorr-diag}
diag(VarCorr(pp_mod)[["Subject"]]) # just the variances
```

We can get the correlation between the intecepts and slopes in two ways. First, by extracting the `"correlation"` attribute and then pulling out the element in row 1 column 2 (`[1, 2]`):

```{r corr1}
attr(VarCorr(pp_mod)[["Subject"]], "correlation")[1, 2] # the correlation
```

Or we can directly compute the value from the variance-covariance matrix itself.

```{r corr2}
# directly compute correlation from variance-covariance matrix
mx <- VarCorr(pp_mod)[["Subject"]]

## if cov = rho * t00 * t11, then
## rho = cov / (t00 * t11).
mx[1, 2] / (sqrt(mx[1, 1]) * sqrt(mx[2, 2]))
```

We can pull out the estimated random effects (BLUPS) using `ranef()`. Like `VarCorr()` , the result is a named list, with each element corresponding to a single random factor.

```{r ranef}
ranef(pp_mod)[["Subject"]]
```

There are other extractor functions that are useful. See `?merMod-class` for details.

We can get fitted values from the model using `fitted()` and residuals using `residuals()`.  (These functions take into account "the conditional modes of the random effects", i.e., the BLUPS).
 
```{r}
mutate(sleep2,
       fit = fitted(pp_mod),
       resid = residuals(pp_mod)) %>%
  group_by(Subject) %>%
  slice(c(1,10)) %>%
  print(n = +Inf)
```

Finally, we can get predictions for new data using `predict()`, as we did above. Below we use `predict()` to imagine what might have happened had we continued our study for three extra days.

```{r extrapolate}
## create the table with new predictor values
ndat <- crossing(Subject = sleep2 %>% pull(Subject) %>% levels() %>% factor(),
                 days_deprived = 8:10) %>%
  mutate(Reaction = predict(pp_mod, newdata = .))
```

```{r extrap-plot, fig.cap="Data against model with extrapolation."}
ggplot(sleep2, aes(x = days_deprived, y = Reaction)) +
  geom_line(data = bind_rows(newdata2, ndat),
            color = 'blue') +
  geom_point() +
  scale_x_continuous(breaks = 0:10) +
  facet_wrap(~Subject) +
  labs(y = "Reaction Time", x = "Days deprived of sleep (0 = baseline)")
```

## Multi-level app

[Try out the multi-level web app](https://shiny.psy.gla.ac.uk/Dale/multilevel){target="_blank"} to sharpen your understanding of the three different approaches to multi-level modeling.

